State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-06-12 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
discussion edits

This paper presents a novel framework for model reduction to facilitate practical state and parameter estimation of spatio-temporal neural fields. A basis function decomposition reduces the infinite dimensional neural field (continuous space), to a finite dimensional state-space model. The level of detail of the estimated field is related to the basis functions decomposition, providing a flexible framework where a trade-off can be made between physiological detail and computational complexity. Using the state-space model, the unscented Rauch Tung Striebel smoother was applied for state estimation. Using the sequence of state estimates, a least squares algorithm was implemented for parameter estimation. Though our novel procedure, we have demonstrated that it is theoretically possible to estimate intracortical connectivity and the synaptic time constant.

The frequency analysis and model selection tools presented in this paper specify the sensor and field basis function arrangement required to form a state-space model that is capable of capturing the dominant cortical dynamics. In addition to providing conditions to ensure the reduced model can capture the significant spectral properties of the field, these new tools can be used to better design electrophysiological recording techniques avoid spatial aliasing. This analysis is becoming increasingly relevant as recording systems are becoming more sophisticated with higher density recordings~\cite{Brinkmann2009}. The effect of the basis function decomposition will restrict the frequency content of the estimated field. This is evident in when comparing Figures~\ref{fig:FieldFFT} and~\ref{fig:EstimatedFieldFFT} where the estimated field is spatially band-limited. The spatial frequency content of the field is governed by disturbance covariance and the shape of the connectivity kernel.
% 
% In practice Even though the some high spatial frequency content of the connectivity kernel is lost through the basis function decomposition, the dominant dynamics are still captured with the estimated reduced model. This is illustrated in Figure~\ref{fig:FieldEstimate}. In addition, the ratio of the kernel parameters follows a tight distribution as seen in Figure~\ref{fig:ParametersRatio}. This indicates that the ratio of excitation to inhibition can be tracked with this approach.

@@

This estimation framework also has theoretical implications by establishing a framework that has the potential to allow testing of hypotheses that have been generated in theoretical studies and to validate neural field models. For example, it has been hypothesised that so-called ``bump solutions'' are a possible mechanism for short term memory formation~\cite{Coombes2005}. Using the proposed model-based framework, we can reconstruct a neural field from data, and estimate parameters and check if they correspond the the theoretically derived parameters where these bump solutions can exist. In addition, the parameters and the state-space can be explored during other cognitive tasks and states of arousal such as during sleep and under anaesthesia.

>>>>

This paper presents a novel framework for model reduction to facilitate practical state and parameter estimation of spatio-temporal neural fields. A basis function decomposition reduces the infinite dimensional neural field (continuous space), to a finite dimensional state-space model. The level of detail of the estimated field is related to the basis functions decomposition, providing a flexible framework where a trade-off can be made between physiological detail and computational complexity. The effect of the basis function decomposition will restrict the frequency content of the estimated field. This is evident in when comparing Figures~\ref{fig:FieldFFT} and~\ref{fig:EstimatedFieldFFT} where the estimated field is spatially band-limited. The spatial frequency content of the field is governed by disturbance covariance and the shape of the connectivity kernel. Using the state-space model, the unscented Rauch Tung Striebel smoother was applied for state estimation. Using the sequence of state estimates, a least squares algorithm was implemented for parameter estimation. Though our novel procedure, we have demonstrated that it is theoretically possible to estimate intracortical connectivity and the synaptic time constant.

The frequency analysis and model selection tools presented in this paper specify the sensor and field basis function arrangement required to form a state-space model that is capable of capturing the dominant cortical dynamics. In addition to providing conditions to ensure the reduced model can capture the significant spectral properties of the field, these new tools can be used to better design electrophysiological recording techniques avoid spatial aliasing. This analysis is becoming increasingly relevant as recording systems are becoming more sophisticated with higher density recordings~\cite{Brinkmann2009}.

% 
% In practice Even though the some high spatial frequency content of the connectivity kernel is lost through the basis function decomposition, the dominant dynamics are still captured with the estimated reduced model. This is illustrated in Figure~\ref{fig:FieldEstimate}. In addition, the ratio of the kernel parameters follows a tight distribution as seen in Figure~\ref{fig:ParametersRatio}. This indicates that the ratio of excitation to inhibition can be tracked with this approach.

@@

This estimation framework also has theoretical implications by establishing a framework that has the potential to allow testing of hypotheses that have been generated in theoretical studies and to validate neural field models. For example, it has been hypothesised that so-called ``bump solutions'' are a possible mechanism for short term memory formation~\cite{Coombes2005}. Using the proposed model-based framework, we can reconstruct a neural field from data, and estimate parameters and check if they correspond the the theoretically derived parameters where these bump solutions can exist. In addition, the parameters and the state-space can be explored during other cognitive tasks and states of arousal such as during sleep and anaesthesia.
